Theorems and Definitions

### 2016-04-07 math_modeling/class/notes/stochastics.tex
number theory

>>>>

\section*{Irreducible Matrices}
We attempted to find the $\pi$ vector in the previous examples. We can use the following theorem to
discuss special Markov Chains.
\begin{quote}
Let $P$ be the transition matrix for an irreducible and aperiodic Markov Chain. Then there exists
a unique probability vector $\pi$ such that
\[
\pi\cdot P=\pi.
\]
\end{quote}
It follows that if $P$ is a regular matrix, then we can also find the $\pi$ vector. A more
interesting phenomenon occurs when we do not have a regular $P$ matrix. A typical model of this can
be seen in the Drunkard's walk without reflection.

\subsection*{The Drunkard's Walk}
The Drunkard's walk represents a random walk with several states. At some states, the drunk man may
either end up in his bed or on the street. In the reflective version of this walk, it is possible that
after going into his bed, the drunkard goes back out. In this example, we want to consider absorbing
states, which are states that are not able to be escaped.

In this walk, we want to model the drunkard by saying that if he falls asleep in the street or he
goes to his bed, then the walk ends. Those conditions are representative of absorbing states.
\[
P=
\left(
\begin{array}{c c c c c}
1 & 0 & 0 & 0 & 0\\
1/2 & 0 & 1/2 & 0 & 0\\
0 & 1/2 & 0 & 1/2 & 0\\
0 & 0 & 1/2 & 0 & 1/2\\
0 & 0 & 0 & 0 & 1
\end{array}
\right).
\]
More generally, an absorbing matrix takes the form of an $r\times r$ matrix with $k$ absorbing
states. It is a fact that ultimately everything will end up in the absorbed state. Markov Chains
are only absorbing when all absorbing state can be reached. If a state is not absorbing, it is
called a transient state.

For the drunkard's walk, we can partition the matrix into four parts:
\[
P=
\left(
\begin{array}{c c c c c}
1 & 0 & 0 & 0 & 0\\
1/2 & 0 & 1/2 & 0 & 0\\
0 & 1/2 & 0 & 1/2 & 0\\
0 & 0 & 1/2 & 0 & 1/2\\
0 & 1/2 & 0 & 1/2 & 0
\end{array}
\right)
=
\left(
\begin{array}{c c c c c}
1 & 0 & 0 & 0 & 0\\
0 & 1 & 0 & 0 & 0\\
1/2 & 0 & 0 & 1/2 & 0\\
0 & 0 & 1/2 & 0 & 1/2\\
0 & 0 & 0 & 0 & 1
\end{array}
\right)
\]
which allows us to calculate a new matrix $N=(I-Q)^{-1}$. Every $n_{ij}$ entry in the $N$ matrix
tells us the average number of times that the process is in the $j$ state if it starts in the $i$
state.

### 2016-04-07 number_theory/hw/hw9.tex
number theory

\item Under what additional assumptions on $m$ and $n$ do we obtain
$\phi(p^{m+n})=\phi(p^{m})\phi(p^{n})$?
\end{enumerate}

\end{enumerate}
\end{document}

>>>>

\item Under what additional assumptions on $m$ and $n$ do we obtain
$\phi(p^{m+n})=\phi(p^{m})\phi(p^{n})$?
\begin{quote}
If either $m$ or $n$, but not both $m$ and $n$ are zero, then we have
$\phi(p^{m+0})=\phi(p^{m})\phi(p^{0})=\phi(p^{m})$ or
$\phi(p^{0+n})=\phi(p^{0})\phi(p^{n})=\phi(p^{n})$.
\end{quote}
\end{enumerate}

\item Suppose that $a$ and $b$ are positive integers.
\begin{enumerate}
\item Prove that $\phi(ab)\geq\phi(a)\phi(b)$.
\begin{quote}
Suppose that $a$ and $b$ are not relatively prime and consider the product $\phi(a)\phi(b)$. Then
\begin{align*}
\phi(a)\phi(b) &<
\phi(p_{1}^{\alpha_{1}}p_{2}^{\alpha_{2}}\dots p_{r}^{\alpha_{r}})
\phi(p_{1}^{\beta_{1}}p_{2}^{\beta_{2}}\dots p_{r}^{\beta_{r}})\\
&= \phi(p_{1}^{\alpha_{1}})\phi(p_{2}^{\alpha_{2}})\dots \phi(p_{r}^{\alpha_{r}})
\phi(p_{1}^{\beta_{1}})\phi(p_{2}^{\beta_{2}})\phi(p_{r}^{\beta_{r}})\\
&= \prod_{i=1}^{r}\left(p_{i}^{\alpha_{i}}-p_{i}^{\alpha_{i}-1}\right)
\prod_{i=1}^{r}\left(p_{i}^{\beta_{i}}-p_{i}^{\beta_{i}-1}\right)\\
&= \prod_{i=1}^{r}p_{i}^{\alpha_{i}}\left(1-{1\over p_{i}}\right)
\prod_{i=1}^{r}p_{i}^{\beta_{i}}\left(1-{1\over p_{i}}\right)\\
&= ab\prod_{i=1}^{r}\left(1-{1\over p_{i}}\right)^{2}.
\end{align*}
We can also consider the function $\phi(ab)$.
\begin{align*}
\phi(ab) &= \phi(p_{1}^{\alpha_{1}}p_{2}^{\alpha_{2}}\dots p_{r}^{\alpha_{r}}
p_{1}^{\beta_{1}}p_{2}^{\beta_{2}}\dots p_{r}^{\beta_{r}})\\
&= \phi(p_{1}^{\alpha_{1}+\beta_{1}}p_{2}^{\alpha_{2}+\beta_{2}}\dots p_{r}^{\alpha_{r}+\beta_{r}})\\
&= \phi(p_{1}^{\alpha_{1}+\beta_{1}})\phi(p_{2}^{\alpha_{2}+\beta_{2}})\dots
\phi(p_{r}^{\alpha_{r}+\beta_{r}})\\
&= \prod_{i=1}^{r}p_{i}^{\alpha_{i}+\beta_{i}}-p_{i}^{\alpha_{i}+\beta_{i}}{1\over p}\\
&= \prod_{i=1}^{r}p_{i}^{\alpha_{i}+\beta_{i}}\left(1-{1\over p_{i}}\right)\\
&= ab\prod_{i=1}^{r}\left(1-{1\over p_{i}}\right).
\end{align*}
Since
\[
\phi(ab)=ab\prod_{i=1}^{r}\left(1-{1\over p_{i}}\right) >
ab\prod_{i=1}^{r}\left(1-{1\over p_{i}}\right)^{2}>\phi(a)\phi(b),
\]
we have
\[
\phi(ab) > \phi(a)\phi(b).
\]
If $gcd(a,b)=1$, then we can say $\phi(ab) = \phi(a)\phi(b)$. Therefore $\phi(ab)\geq\phi(a)\phi(b)$.
\end{quote}

\item Prove that $\phi(ab) = \phi(a)\phi(b)$ if and only if $gcd(a,b)=1$.
\begin{quote}
$\Rightarrow)$ Suppose $\phi(ab) = \phi(a)\phi(b)$ implies $gcd(a,b)>1$. We know from $(a)$ that
when $a$ and $b$ are not relatively prime, $\phi(ab) > \phi(a)\phi(b)$, which is a direct contradiction
to $\phi(ab) = \phi(a)\phi(b)$. Therefore $\phi(ab) = \phi(a)\phi(b)$ implies $gcd(a,b)=1$.
\noindent\\\\
$\Leftarrow)$ Suppose $gcd(a,b)=1$. Then $a$ and $b$ are relatively prime, so
$\phi(ab)=\phi(a)\phi(b)$ by the theorem.
\end{quote}
\end{enumerate}

\item Suppose that $n$ is a positive integer and $k$ is any integer.
\begin{enumerate}
\item Prove that $gcd(n,k)=1$ if and only if $gcd(n,n-k)=1$.
\begin{quote}
$\Leftarrow)$ Suppose that $gcd(n,k)>1$. Then $gcd(n,k)=q$, so $q\mid n$ and
$q\mid k$, $n=qa$ and $k=qb$ for some $a,b\in\ZZ$. Then $n-k=pa-pb=p(a-b)$, so $p\mid n-k$, which
implies that $gcd(n,n-k)>1$. Therefore, by contrapositive, $gcd(n,k)=1$ implies $gcd(n,n-k)=1$.
\noindent\\\\
$\Rightarrow)$ Suppose $gcd(n,n-k)>1$. Then there exists $p\in\ZZ$ such that $p\mid n$ and $p\mid k$.
Then $n=pa$ and $k=pb$ for some $a,b\in\ZZ$. We can express $n-k=pa-pb=p(a-b)$, so $p\mid(n-k)$ and
$p\mid n$, and $gcd(n,n-k)>1$.
\end{quote}
\item Prove that $\phi(n)$ is an even integer for all $n\geq3$.
\begin{quote}
By definition, the totient function counts the number of units in $\ZZ_{n}$. Suppose that
$n\geq3$. If we take any element $k$ from $\ZZ_{n}$, we know by $(a)$ that if $k$ is relatively
prime to $n$, then $n-k$ is relatively prime to $n$. Therefore, if $k$ is a unit in $\ZZ_{n}$, then
$n-k$ is also a unit in $\ZZ_{n}$. Then for every unit $k$ in $\ZZ_{n}$, we can find another unit
$n-k$ also in $\ZZ_{n}$, which means there are an even number of units in $\ZZ_{n}$, so
$\phi(n)\in\{ 2j\ |\ j\in\ZZ \}$.
\end{quote}
\end{enumerate}

\item If $n$ is a positive integer prove that $\phi(n)=2$ if and only if $n\in\{ 3,4,6\}$.
\begin{quote}
$\Rightarrow)$
\noindent\\\\
$\Leftarrow)$ Suppose $n\in\{ 3,4,6\}$. Then $\phi(3)=2$, $\phi(4)=2$, and $\phi(6)=2$.
\end{quote}
\end{enumerate}
\end{document}
